Distortion-corrected kernel density estimate on Riemannian manifolds

### 2022-07-19 paper/paper.Rmd
Sec 2.4 rewrite

For $p \in M$, denote an open ball centered at $p$ with radius $r$ as $B(p,r)$ and the open neighborhood of $0_p \in T_pM$ can be denoted as $B(0_p,r)=\exp_p^{-1}(B(p,r))$. Consider the Riemannian metrics $g^{\prime}$ and $g^{\prime \prime}$ in $B(0_p,r)$, with $g^{\prime} = \exp^{\ast}_p(g)$ being the pullback of $g$ induced by the exponential map and $g^{\prime \prime}$ being the canonical metric induced by $g_p$ in $B(0_p,r)$. 
Let $u \in B(0_p,r)$ be a tangent vector and for any given chart $(\bar{U}, \bar{\psi})$ of $B(0_p,r)$, we have $u \in \bar{U}$. 
The volumes of the parallelepiped spanned by the basis $\{(\partial / \partial\bar{\psi}_1|_u), \dots, (\partial / \partial\bar{\psi}_d|_u)\}$ are given by $| \det g^{\prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2}$ with respect to $g^{\prime}$, and $| \det g^{\prime \prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2}$ with respect to $g^{\prime \prime}$. The quotient of these two volumes is then independent of the chart $(\bar{U}, \bar{\psi})$ [@Henry2009-om]. 
Therefore, given a point $q \in B(p,r)$ and a tangent vector $u = \exp^{-1}_p(q) \in B(0_p,r)$, for any chart $(\bar{U}, \bar{\psi})$ of $B(0_p,r)$ that contains $u$, @Henry2009-om defines the volume density function on the Riemannian manifold $(M,g)$ as
$$
\theta_p(q) = \frac{| \det g^{\prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2} }{ | \det g^{\prime \prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2} } \\ .
$$
The volume density function is only defined for points satisfying $d_g(p,q) < inj_gM$ so that there exists an open ball $B(p,r) \in T_pM$. And $\theta_p(q)=1$ for all $p,q \in \mathbb{R}^d$ when the manifold $M$ is $\mathbb{R}^d$ with the canonical metric.

Now we consider the case with the manifold learning embedding. Consider the chart $(U, \psi)$ induced by an orthonormal basis of the tangent space of the embedding $T_{\pmb{y}_i}E$ and a normal neighborhood of $\pmb{y}_i$. Given a query point $\pmb{p} \in B(\pmb{y}_i,r)$, the volume of the parallelepiped is spanned by the same basis.  
Then the volume density function is given by 
$$
\theta_{\pmb{y}_i}(\pmb{p}) = \frac{|\det \pmb{H}(\pmb{p})|^{1/2}}{|\det \pmb{H}(\pmb{y}_i)|^{1/2}},
$$ 
where $\pmb{H}$ is the estimated Riemannian metric for the unknown manifold.
Again, the volume density function is not defined for all $\pmb{p}$ but for points within the geodesic ball satisfying $\pmb{p} \in B(\pmb{y}_i,r)$, which is related to the support of $K$ being $[0,1]$ in condition (v) of \@ref(eq:kernelcondition).

@@

As pointed out by @Perrault-Joncas2013-pq, the naive distance $\|p-q\|$ underestimates $d_g$, and the graph shortest distance $d_G$ overestimates $d_g$ by the triangle inequality. By using only the first-order correction in $\hat{d}_g$, our estimator is also affected by the estimated Riemannian metric. In order to correct for the estimation error in $\pmb{H}(\pmb{y}_i)$, we further divide the density estimator in \@ref(eq:denestimator) by a constant term $C$ so that the integral of the kernel is equal to 1. The constant term in $1$-d is given by the *area under the curve (AUC)*, and can be named as the *volume between the surface (VBS)* in higher dimension. We use the volume estimator in Section 6.4 of @Perrault-Joncas2013-pq to sum over all embedding points $\pmb{y}_i$ and get
$$
C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i),
$$
where the distortion is corrected for the tangent plane of the embedding points.

@@

An important goal of manifold learning is to recover the local or global features of the data. In order to recover the geometry of the manifold, @Perrault-Joncas2013-pq propose a method to augment any existing reasonable embedding and allow for the computation of geometric values to be calculated with an estimation of the Riemannian metric $g$. This could also be used to fix the geometric distortion in the embedding. It is worth noticing that the Riemannian metric from @Perrault-Joncas2013-pq is a $d\times d$ positive definite matrix for each data point. Therefore, we propose to take the geometric distortion into consideration when estimating the kernel densities of the embedding with Equation \@ref(eq:denestimator). In this way, we present a consistent kernel density estimator with pointwise variable bandwidth as the Riemannian metric.

The Riemannian metric $g$ is a symmetric and positive definite tensor field which defines an inner product $\langle,\rangle_g$ on the tangent space $T_pM$ for every point $p \in M$. If the inner product of the tangent space is known for a given geometry, the Riemannian metric is a good measure to recover the geometry of the manifold. The Metric Learning algorithm [@Perrault-Joncas2013-pq] then augment the embedded manifold with the Riemannian metric and produce a Riemannian manifold $(M, g)$.

To recover the original geometry of the manifold, we need to know what the inner product corresponds to in the embedding. The inner product between two vectors $u,v \in T_pM$, $\langle u,v \rangle_g=g_{ij}u^iv^j$[^1], can be used to define some geometric quantities, such as the vector norm $\|u\|=\sqrt{\langle u,v \rangle_g}$ and the angle between two vectors $\cos{\theta}=\frac{\langle u,v \rangle_g}{\|u\|\|v\|}$ in the tangent space.
Therefore, for each point $p\in M$ in any coordinate system, the Riemannian metric $g$ is a $d\times d$ symmetric positive definite matrix, where $d$ is the dimension of the manifold.

[^1]: Here the Einstein notation is used where superscripts denote summation over $i$ and $j$

The line element and volume element of the full manifold or a subset of the manifold can also be computed from $g$. The arc length of a curve $c\in M$ is defined as 
$$
l(c)=\int_a^b \sqrt{g_{ij} \frac{dx^i}{dt} \frac{dx^j}{dt}} dt,
$$ 
where $(x^1,\dots,x^d)$ are the coordinates of chart $(U,x)$ and $c(t)$ is a function mapping $[a,b]$ to $M$. While the volume of $V\subset M$ is computed by 
$$
Vol(V)=\int_V \sqrt{\|g\|} dx^1\dots dx^d.
$$ 
Both the concepts of distance and volume are relevant to kernel density estimation.

@Perrault-Joncas2013-pq propose the Learn metric algorithm which mainly involves four steps: weighted neighborhood graph construction; geometric graph Laplacian calculation; manifold learning; and Riemannian metric calculation. We restate these four steps in Algorithm \@ref(alg:learnmetric). 
By applying an existing manifold learning algorithm to the data $X\in \mathbb{R}^r$ with $n$ observations, a low-dimensional embedding $f_n \in \mathbb{R}^d$ can be computed. Most manifold learning methods involve the construction of the nearest
neighbor graph based on which the Laplace-Beltrami operator $\Delta_M$ is built. The Laplacian is quite useful because it can be coordinate-free while containing all the important geometry.
@Perrault-Joncas2013-pq have stated one way to compute the approximated $\Delta_M$ with a discrete consistent estimator, the geometric graph Laplacian $\mathcal{L}_{\varepsilon,n}$ [@Zhou2011-za], where $\varepsilon$ is the radius parameter for the nearest neighbor graph.
The graph Laplacian together with the embedding can be used in the Metric Learning algorithm to achieve the augmented embedding with the Riemannian metric $(f_n, g_n)$.

As pointed out by @Perrault-Joncas2013-pq, if the embedding dimension $s$ is larger than the manifold intrinsic dimension $d$, the rank of the embedding metric $h_n(p)$ is $d$; otherwise, the Riemannian metric $g_n$ will be returned. This algorithm is also implemented in a Python library *megaman* [@McQueen2016-xz]. It is designed to apply the manifold learning methods to large-scale data sets, as well as compute the Riemannian metric of the manifold.

<!-- ```{=tex} -->
\begin{algorithm}[!htb]
  \caption{Learn metric algorithm in \cite{Perrault-Joncas2013-pq} }
  \label{alg:learnmetric}
  \DontPrintSemicolon
  \SetAlgoLined
  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}\SetKwInOut{OptParameter}{optimization parameter}
  \Input{ high-dimensional data $x_i \in \mathcal{R}^s$ for all $i=1,\ldots,N$ }
  \Output{ low-dimensional data $y_i \in \mathcal{R}^d$ and its Riemannian metric $h_i$ for all $i=1,\ldots,N$ }
  \Parameter{ embedding dimension $d$, bandwidth parameter $\sqrt{\varepsilon}$, manifold learning algorithm }
  \OptParameter{ manifold learning parameters }
  \BlankLine
  \begin{algorithmic}[1]

  \STATE Construct a weighted neighborhood graph $\mathcal{G}_{w,\varepsilon}$ with weight matrix $W$ where $w_{i,j}=\exp(-\frac{1}{\varepsilon}\|x_i-x_j\|^2)$ for data points $x_i,x_j \in \mathbb{R}^s$;

@@

  \STATE Embed each data point $x_i\in \mathcal{R}^s$ to embedding coordinates $\pmb{y}=(\pmb{y}^1,\dots,\pmb{y}^d)^\prime$ by any existing manifold learning algorithm;

  \STATE Obtain the matrix $\pmb{\tilde{h}}$ of all data point by applying the graph Laplacian $\widetilde{\mathcal{L}}_{\sqrt{\varepsilon},N}$ to the embedding coordinates matrix $\pmb{y}$ with each element vector in $\pmb{\tilde{h}}$ being
  $$
    \pmb{\tilde{h}}^{i j} = \frac{1}{2} \left[\tilde{\mathcal{L}}_{\varepsilon, N}\left(y^i \cdot \pmb{y}^j\right) - \pmb{y}_i \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^j\right) - \pmb{y}^j \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^i\right)\right],
  $$
  where $i,j=1,\dots,d$ and the $\cdot$ calculation is the elementwise product between two vectors;

  \STATE Calculate the Riemannian metric $\pmb{h}$ as the rank $d$ pseudo inverse of $\tilde{\pmb{    h}}$ with 
  $$
    \pmb{h} = U diag{1/(\Lambda[1:d])} U^\prime,
  $$
  where $[U, \Lambda]$ is the eigendecomposition of matrix $\pmb{\tilde{h}}(x)$, and $U$ is the matrix of column eigenvectors ordered by the eigenvalues $\Lambda$ in descending order.

@@

>>>>

For $p \in M$, denote an open ball centered at $p$ with radius $r$ as $B(p,r)$ and the open neighborhood of $0_p \in T_pM$ can be denoted as $B(0_p,r)=\exp_p^{-1}(B(p,r))$. Consider the Riemannian metrics $g^{\prime}$ and $g^{\prime \prime}$ in $B(0_p,r)$, with $g^{\prime} = \exp^{\ast}_p(g)$ being the pullback of $g$ induced by the exponential map and $g^{\prime \prime}$ being the canonical metric induced by $g_p$ in $B(0_p,r)$. 
Let $u \in B(0_p,r)$ be a tangent vector and for any given chart $(\bar{U}, \bar{\psi})$ of $B(0_p,r)$, we have $u \in \bar{U}$. 
The volumes of the parallelepiped spanned by the basis $\{(\partial / \partial\bar{\psi}_1|_u), \dots, (\partial / \partial\bar{\psi}_d|_u)\}$ are given by $| \det g^{\prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2}$ with respect to $g^{\prime}$, and $| \det g^{\prime \prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2}$ with respect to $g^{\prime \prime}$. The quotient of these two volumes is then independent of the chart $(\bar{U}, \bar{\psi})$ [@Henry2009-om]. 
Therefore, given a point $q \in B(p,r)$ and a tangent vector $u = \exp^{-1}_p(q) \in B(0_p,r)$, for any chart $(\bar{U}, \bar{\psi})$ of $B(0_p,r)$ that contains $u$, @Henry2009-om defines the volume density function on the Riemannian manifold $(M,g)$ as
\begin{equation}
\label{eq:volumedensityratio}
\theta_p(q) = \frac{| \det g^{\prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2} }{ | \det g^{\prime \prime}((\partial / \partial\bar{\psi}_i|_u), (\partial / \partial\bar{\psi}_j|_u)) |^{1/2} } \\ .
\end{equation}
The volume density function is only defined for points satisfying $d_g(p,q) < inj_gM$ so that there exists an open ball $B(p,r) \in T_pM$. And $\theta_p(q)=1$ for all $p,q \in \mathbb{R}^d$ when the manifold $M$ is $\mathbb{R}^d$ with the canonical metric.

Now we consider the case with the manifold learning embedding. Consider the chart $(U, \psi)$ induced by an orthonormal basis of the tangent space of the embedding $T_{\pmb{y}_i}E$ and a normal neighborhood of $\pmb{y}_i$. Given a query point $\pmb{p} \in B(\pmb{y}_i,r)$, the volume of the parallelepiped is spanned by the same basis.  
Then the volume density function is given by 
\begin{equation}
\label{eq:volumedensityest}
\theta_{\pmb{y}_i}(\pmb{p}) = \frac{|\det \pmb{H}(\pmb{p})|^{1/2}}{|\det \pmb{H}(\pmb{y}_i)|^{1/2}},
\end{equation} 
where $\pmb{H}$ is the estimated Riemannian metric for the unknown manifold.
Again, the volume density function is not defined for all $\pmb{p}$ but for points within the geodesic ball satisfying $\pmb{p} \in B(\pmb{y}_i,r)$, which is related to the support of $K$ being $[0,1]$ in condition (v) of \@ref(eq:kernelcondition).

@@

As pointed out by @Perrault-Joncas2013-pq, the naive distance $\|p-q\|$ underestimates $d_g$, and the graph shortest distance $d_G$ overestimates $d_g$ by the triangle inequality. By using only the first-order correction in $\hat{d}_g$, our estimator is also affected by the estimated Riemannian metric. In order to correct for the estimation error in $\pmb{H}(\pmb{y}_i)$, we further divide the density estimator in \@ref(eq:denestimator) by a constant term $C$ so that the integral of the kernel is equal to 1. The constant term in $1$-d is given by the *area under the curve (AUC)*, and can be named as the *volume between the surface (VBS)* in higher dimension. We use the volume estimator in Section 6.4 of @Perrault-Joncas2013-pq to sum over all embedding points $\pmb{y}_i$ and get
\begin{equation}
\label{eq:vbs}
C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i),
\end{equation}
where the distortion is corrected for the tangent plane of the embedding points.

@@

An important goal of manifold learning is to recover the local or global features of the data. However, manifold learning algorithms usually fail to generate an isometric embedding.
In order to recover the geometry of the manifold, @Perrault-Joncas2013-pq propose a method to augment any existing reasonable embedding and allow for the computation of geometric values to be calculated with an estimated Riemannian metric. Instead of developing a new manifold learning method, they aim to fix the geometric distortion in the embedding. It is worth noticing that the Riemannian metric from @Perrault-Joncas2013-pq is a $d\times d$ positive definite matrix variable at each data point. The estimated Riemannian metric completes our proposed kernel density estimation in \autoref{dckde} and enables us to take the geometric distortion into consideration when estimating the densities of the manifold.

The Riemannian metric $g$ is a symmetric and positive definite tensor field which defines an inner product $\langle,\rangle_g$ on the tangent space $T_pM$ for every point $p \in M$. If the inner product of the tangent space is known for a given geometry, the Riemannian metric is a good measure to recover the geometry of the manifold. For each point $p\in M$ in any embedding coordinate system, the Riemannian metric $g$ is a $d\times d$ symmetric positive definite matrix, where $d$ is the dimension of the manifold.

The inner product between two vectors $u,v \in T_pM$, given by $\langle u,v \rangle_g=g_{ij}u^iv^j$[^1], can be used to define some geometric quantities, such as the vector norm $|u|=\sqrt{\langle u,u \rangle_g}$ and the angle between two vectors $\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}$ in the tangent space.
[^1]: Here the Einstein notation is used where superscripts denote summation over $i$ and $j$.
The line element and volume element of the full manifold or a subset of the manifold can also be computed from $g$. The arc length of a curve $c$ along the manifold is defined as
$$
l(c)=\int_a^b \sqrt{g_{ij} \frac{dx^i}{dt} \frac{dx^j}{dt}} dt,
$$ 
where $(x^1,\dots,x^d)$ are the coordinates of chart $(U,x)$ and $c(t)$ is a function that maps $[a,b]$ to $M$. While the volume of $V\subset M$ is computed by 
$$
Vol(V)=\int_V \sqrt{|\det g|} dx^1\dots dx^d.
$$ 
Both the concepts of distance and volume are relevant to kernel density estimation as we have shown in Equation \@ref(eq:volumedensityratio), \@ref(eq:volumedensityest), \@ref(eq:dgestimate), and \@ref(eq:vbs).

@Perrault-Joncas2013-pq propose the Learn Metric algorithm to augment any manifold learning embedding with the estimated Riemannian metric. There are mainly four steps in the algorithm: weighted neighborhood graph construction; geometric graph Laplacian calculation; manifold learning; and Riemannian metric calculation. We restate these four steps in Algorithm \@ref(alg:learnmetric). 
By applying an existing manifold learning algorithm to the high-dimensional input data $\pmb{X}\in \mathbb{R}^s$ with $N$ observations, a low-dimensional embedding $\pmb{Y} \in \mathbb{R}^d$ can be computed. Most manifold learning methods involve the construction of the nearest neighbor graph based on which the Laplace-Beltrami operator $\Delta_M$ is built. The Laplacian is quite useful because it can be coordinate-free while containing all the important geometry.
@Perrault-Joncas2013-pq have stated one way to compute the approximated $\Delta_M$ with a discrete consistent estimator, the geometric graph Laplacian $\hat{\mathcal{L}}_{\varepsilon,n}$ [@Zhou2011-za], where $\varepsilon$ is the radius parameter for the nearest neighbor graph.
The graph Laplacian estimator together with the embedding $\pmb{Y}$ could further be used to achieve the augmented embedding with the Riemannian metric, $(\pmb{Y}, \pmb{H})$.
This algorithm is implemented in a Python library *megaman* [@McQueen2016-xz]. It is designed to apply the manifold learning methods to large-scale data sets, as well as compute the Riemannian metric of the manifold. We have re-implemented the algorithm in *R*.

As pointed out by @Perrault-Joncas2013-pq, if the embedding dimension is larger than the manifold intrinsic dimension $d$, the rank of the embedding metric $\pmb{H}(p)$ is equal to $d$; otherwise, the Riemannian metric will be returned. 
To ensure that the estimated Riemannian metric is returned as a $d\times d$ matrix, for unknown manifold dimension, we apply the intrinsic dimension (id) selection algorithm called *two-nearest neighbor estimator (TWO-NN estimator)* [@Facco2017-rl; @Denti2021-jl] to derive id estimator from a Poisson point process. @Facco2017-rl point out that of the distances between a point and its second and first neareast neighbors follow a Pareto distribution with scale parameter $1$ and shape parameter $d$, the intrinsic dimension of the manifold [@Facco2017-rl]. By fitting a Pareto distribution to proper transformed data, the intrinsic dimension could be estimated. An *R* library named *intRinsic* has implemented this algorithm and we are going to apply it to when $d$ is unknown.

\begin{algorithm}[!htb]
  \caption{Learn metric algorithm in \cite{Perrault-Joncas2013-pq} }
  \label{alg:learnmetric}
  \DontPrintSemicolon
  \SetAlgoLined
  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}\SetKwInOut{OptParameter}{optimization parameter}
  \Input{ high-dimensional data $\pmb{x}_i \in \mathbb{R}^s$ for all $i=1,\ldots,N$ }
  \Output{ low-dimensional data $\pmb{y}_i \in \mathbb{R}^d$ and its Riemannian metric $\pmb{H}(\pmb{y}_i)$ for all $i=1,\ldots,N$ }
  \Parameter{ embedding dimension $d$, bandwidth parameter $\sqrt{\varepsilon}$, manifold learning algorithm }
  \OptParameter{ manifold learning parameters $MANILEARN$}
  \BlankLine
  \begin{algorithmic}[1]

  \STATE Construct a weighted neighborhood graph $\mathcal{G}_{w,\varepsilon}$ with weight matrix $\pmb{W}$ where $w_{i,j}=\exp(-\frac{1}{\varepsilon}\|\pmb{x}_i-\pmb{x}_j\|^2)$ for data points $\pmb{x}_i,\pmb{x}_j \in \mathbb{R}^s$;

@@

  \STATE Embed all data point $\pmb{X}\in \mathbb{R}^s$ to embedding coordinates $\pmb{Y}=(\pmb{y}^1,\dots,\pmb{y}^d)^\prime$ by any existing manifold learning algorithm $MANILEARN$;

  \STATE Obtain the matrix $\pmb{\tilde{H}}$ of all data point by applying the graph Laplacian $\widetilde{\mathcal{L}}_{\sqrt{\varepsilon},N}$ to the embedding coordinates matrix $\pmb{Y}$ with each element vector in $\pmb{\tilde{H}}$ being
  $$
    \pmb{\tilde{H}}^{i j} = \frac{1}{2} \left[\tilde{\mathcal{L}}_{\varepsilon, N}\left(\pmb{y}^i \cdot \pmb{y}^j\right) - \pmb{y}_i \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^j\right) - \pmb{y}^j \cdot\left(\tilde{\mathcal{L}}_{\varepsilon, n} \pmb{y}^i\right)\right],
  $$
  where $i,j=1,\dots,d$ and the $\cdot$ calculation is the elementwise product between two vectors;

  \STATE Calculate the Riemannian metric $\pmb{H}$ as the rank $d$ pseudo inverse of $\tilde{\pmb{H}}$ with 
  $$
    \pmb{H} = \pmb{U} diag{1/(\Lambda[1:d])} \pmb{U}^\prime,
  $$
  where $[\pmb{U}, \Lambda]$ is the eigendecomposition of matrix $\pmb{\tilde{h}}(x)$, and $U$ is the matrix of column eigenvectors ordered by the eigenvalues $\Lambda$ in descending order.

@@

Now we are able to generate kernel density estimates of the underlying manifold using the estimator in Equation \@ref(eq:denestimator).
